A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-03 shanghai.tex
Edits, mostly in section 8

 \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} : \gqcs{L}_{\Frob{\Fq}(x)} \to \gqcs{L}_x$
  such that
  \[
   \begin{tikzcd}[row sep=40]
    \gqcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
    && \gqcs{L}_{\Frob{\Fq}(x)}\otimes\gqcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
    \gqcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
    && \gqcs{L}_x \otimes\gqcs{L}_y
   \end{tikzcd}
  \]
  commutes, for all $x,y\in G(\bFq)$.
\end{enumerate}
Under this equivalence, a morphism $\alpha : \qcs{L} \to \qcs{L'}$ of quasicharacter sheaves on $G$ is given by 
\begin{enumerate}
 \labitem{(cs.3)}{cs.3} an indexed set $\alpha_x : \gqcs{L}_x \to \gqcs{L'}_x$
  of linear transformations (of one-dimensional $\EE$-vector spaces),
  as $x$ runs over $G(\bFq)$, such that
  \[
   \begin{tikzcd}[column sep=40]
    \arrow[swap]{d}{\phi_x} \gqcs{L}_{\Frob{\Fq}(x)} \rar{\alpha_{\Frob{\Fq}(x)}} & \gqcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
    &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}
    & \arrow[swap]{d}{\mu} \gqcs{L}_{x+y} \rar{\alpha_{x+y}} & \gqcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
    \gqcs{L}_x \rar{\alpha_x} & \gqcs{L'}_x
    & {} & \gqcs{L}_x\otimes\gqcs{L}_y \rar{\alpha_x\otimes\alpha_y} & \gqcs{L'}_x \otimes\gqcs{L'}_y
   \end{tikzcd}
  \]
  both commute.
\end{enumerate}

\begin{lemma}\label{lemma:section}
 If $\qcs{L}$ is a quasicharacter sheaf on a commutative, \'etale group scheme $G$
 over $\Fq$ and if $G(\bFq)$ is finitely generated, then there is a global section
 $s \ceq \{ s(x) \in \gqcs{L}_x \tq x \in G(\bFq) \}$ satisfying
 \begin{equation}\label{section}
 \end{equation}
\end{lemma}

@@

  There is a unique isomorphism $\gqcs{L}_0 \cong \EE$ under which $\mu_{0,0}$ corresponds
  to the multiplication map $\EE \otimes\EE \to \EE$.
  The condition $\mu_{0,0}(s(0)) = s(0) \otimes s(0)$ together with the non-triviality of $s$
  then forces $s(0) = 1$.

@@

\begin{proof}
  Suppose $\qcs{L}$ is a quasicharacter sheaf on $G$.  
  Using Lemma~\ref{lemma:section}, 
  we may choose a section $s \ceq \{ s(x) \in \gqcs{L}_x \tq x \in G(\bFq) \}$ satisfying \eqref{section}.
  We will construct a group isomorphism between $\QCiso{G}$ and
  $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$ using $s$, then prove independence from the choice of $s$.

@@

  Although we used a global section $s$ in the definition of
  $\tau_\qcs{L}$, the class of $\tau_\qcs{L}$ does not depend on this
  choice. To see why, let $s_1$ and $s_2$ be two global sections of
  $\qcs{L}$ and let $\tau_1$ and $\tau_2$ be the cocycles defined by $s_1$ and $s_2$.
  For each $x \in G(\bFq)$ there is a unique scalar
  $a(x) \in \EEx$ such that $s_2(x) = a(x) s_1(x)$. Then \eqref{section} and
  condition~\ref{cs.1} together imply $a(x+y) = a(x)a(y)$, so
  $a\in G(\bFq)^*$.  The coboundary $\tau_0 : \Weil{\Fq} \to G(\bFq)^*$
  defined by $\tau_0(\Frob{\Fq}) \ceq \,^{\Frob{\Fq}} a \cdot a^{-1}$ satisfies $\tau_2 = \tau_0 \tau_1$,
  so $\tau_2$ and $\tau_1$ are cohomologous. Thus
  \begin{equation}\label{cohomologous}
  \begin{split}
    \obj \QC(G) &\to \Hh^1(\Weil{\Fq},G(\bFq)^*), \\
    \qcs{L} &\mapsto [\tau_\qcs{L}],
  \end{split}
  \end{equation}
  is independent of the choice of global sections made above.

  We now show that $[\tau_\qcs{L}]$ depends only on the isomorphism class of $\qcs{L}$.
  Let $\alpha : \qcs{L} \to \qcs{L'}$ be an isomorphism and
  let $\tau$ and $\tau'$ be any cocycles representing the classes in the
  image of $\qcs{L}$ and $\qcs{L'}$ under \eqref{cohomologous}; by construction,
  $\tau$ and $\tau'$ are defined by the choice of sections $s$ and $s'$ 
  according to Lemma~\ref{lemma:section}.
    Now, $\varphi_x(s(x))$ is a
  non-zero scalar multiple of $s'(x)$; define $a: G(\bFq) \to \EEx$
  by $\alpha_x(s(x)) = a(x) s'(x)$.
  By the second part of condition~\ref{cs.3},
  $a(x+y) = a(x)a(y)$, so $a \in G(\bFq)^*$. By the first part of
  condition~\ref{cs.3}, $\tau'(\Frob{\Fq}) = \tau(\Frob{\Fq}) \cdot (\,^{\Frob{\Fq}^{-1}}a \cdot a^{-1})$,
  %\Clifton{Fix this.}
  so $\tau$ and $\tau'$ are cohomologous. This shows that
  \begin{equation}\label{pich1}
  \begin{split}
    \QCiso{G} &\to \Hh^1(\Weil{\Fq},G(\bFq)^*),\\
    [\qcs{L}] &\mapsto [\tau_\qcs{L}]
  \end{split}
  \end{equation}
  is well-defined.

@@

 To see that \eqref{pich1} is injective,
 suppose $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$.
 Let $s$ (resp. $s'$) be a section appearing in the definition of $\tau_\qcs{L}$ (resp. $\tau_\qcs{L'}$);
 recall that $s$ (resp. $s'$) is compatible with $\mu$ (resp $\mu'$) 
 in the sense of Lemma~\ref{lemma:section}.
 Since $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$, 
 it follows that $\tau_\qcs{L}$ and $\tau_{\qcs{L'}}$ are cohomologous, 
  so there is some $a\in G(\Fq)^*$ such that 
  $\tau'(\Frob{\Fq}) = \tau(\Frob{\Fq}) \cdot (\,^{\Frob{\Fq}}a \cdot a^{-1})$. 
  Define $\alpha : \qcs{L}\to \qcs{L'}$ by $\alpha_x(s(x)) \ceq a(x) s'(x)$, for each $x\in G(\bFq)$.
  It is straightforward to check that $\alpha$ satisfies condition~\ref{cs.3}
  and that $\alpha^{-1} : \qcs{L'}\to \qcs{L}$, defined by $\alpha^{-1}_x(s'(x)) \ceq a(x)^{-1} s(x)$, 
  is its inverse. Thus, $\qcs{L} \iso \qcs{L'}$ in $\QC(G)$. 
  This shows that \eqref{pich1} is injective.

 Finally, to show that \eqref{pich1} is surjective, pick $[\tau]\in \Hh^1(\Weil{\Fq}, G(\bFq)^*)$.
 For each $x\in G(\bFq)$, set $\gqcs{L}_x \ceq \EE$.
 Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by
 $\mu_{x,y}(1) \ceq 1\otimes1$, for $x,y\in G(\bFq)$; 
 clearly, $\mu$ satisfies condition~\ref{cs.1}.
 Define $\phi_x : \gqcs{L}_{\Frob{\Fq}} \to \gqcs{L}_x$ by $\phi_x(1) \ceq \tau(\Frob{\Fq})(x)$,
 for each $x\in G(\bFq)$. 
 Then $\phi$ satisfies condition~\ref{cs.2} because $\tau \in Z^1(\Weil{\Fq},G(\bFq)^*)$.
 In this way we have built a quasicharacter sheaf $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ 
 from $\tau \in \Hh^1(\Weil{\Fq}, G(\bFq)^*)$ such that the image of $\qcs{L}_{/\text{iso}}$
 under \eqref{pich1} is the class $[\tau]$ in $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$.
 This shows that \eqref{pich1} is surjective, 
 and completes the proof that it is an isomorphism of groups.
\end{proof}

@@

Recall from Section~\ref{sec:finite} that 
a morphism $f : H\to G$ of group $\Fq$-schemes is a discrete isogeny 
if it is a finite, surjective \'etale morphism and 
the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $f^{-1}({\bar e})$ is trivial.

>>>>

 \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} : \gqcs{L}_{\Frob{\Fq}(x)} \to \gqcs{L}_x$
  such that
  \[
   \begin{tikzcd}[row sep=40]
    \gqcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
    && \gqcs{L}_{\Frob{\Fq}(x)}\otimes\gqcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
    \gqcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
    && \gqcs{L}_x \otimes\gqcs{L}_y
   \end{tikzcd}
  \]
  commutes, for all $x,y\in G(\bFq)$.
\end{enumerate}
Under this equivalence, a morphism $\alpha : \qcs{L} \to \qcs{L'}$ of quasicharacter sheaves on $G$ is given by 
\begin{enumerate}
 \labitem{(cs.3)}{cs.3} an indexed set $\alpha_x : \gqcs{L}_x \to \gqcs{L'}_x$
  of linear transformations such that
  \[
   \begin{tikzcd}[column sep=40]
    \arrow[swap]{d}{\phi_x} \gqcs{L}_{\Frob{\Fq}(x)} \rar{\alpha_{\Frob{\Fq}(x)}} & \gqcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
    &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}
    & \arrow[swap]{d}{\mu} \gqcs{L}_{x+y} \rar{\alpha_{x+y}} & \gqcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
    \gqcs{L}_x \rar{\alpha_x} & \gqcs{L'}_x
    & {} & \gqcs{L}_x\otimes\gqcs{L}_y \rar{\alpha_x\otimes\alpha_y} & \gqcs{L'}_x \otimes\gqcs{L'}_y
   \end{tikzcd}
  \]
  both commute for all $x, y \in G(\bFq)$.
\end{enumerate}

A \emph{global section} of $\qcs{L}$ is a function $s : G(\bFq) \to \coprod_{x \in G(\bFq)} \gqcs{L}_x$ so that $s(x) \in \gqcs{L}_x$ for all $x \in G(\bFq)$ and
\begin{equation}\label{section}
\mu_{x,y}(s(x+y)) = s(x) \otimes s(y).
\end{equation}

We say that a section is \emph{trivial} if $s(x) = 0$ for all $x \in G(\bFq)$ and \emph{nontrivial} if $s(x) \ne 0$ for all $x \in G(\bFq)$.

\begin{lemma}\label{lemma:section}
 If $\qcs{L}$ is a quasicharacter sheaf on a commutative, \'etale group scheme $G$
 over $\Fq$ and if $G(\bFq)$ is finitely generated, then $\qcs{L}$ has a nontrivial global section.
\end{lemma}

@@

  There is a unique isomorphism $\gqcs{L}_0 \cong \EE$ under which $\mu_{0,0}^{-1}$ corresponds
  to the multiplication map $\EE \otimes\EE \to \EE$.
  The condition $\mu_{0,0}(s(0)) = s(0) \otimes s(0)$ together with the non-triviality of $s$
  then forces $s(0) = 1$.

@@

\begin{proof}
  Suppose $\qcs{L}$ is a quasicharacter sheaf on $G$ and choose a global section $s$
  of $\qcs{L}$ using Lemma~\ref{lemma:section}.
  We will construct a group isomorphism between $\QCiso{G}$ and
  $\Hh^1(\Weil{\Fq}, G(\bFq)^*)$ using $s$, then prove independence from the choice of $s$.

@@

  Although we used a global section $s$ in the definition of
  $\tau_\qcs{L}$, the class of $\tau_\qcs{L}$ does not depend on this
  choice. To see why, let $s_1$ and $s_2$ be two nontrivial global sections of
  $\qcs{L}$ and let $\tau_1$ and $\tau_2$ be the cocycles defined by $s_1$ and $s_2$.
  For each $x \in G(\bFq)$ set $a(x) = \frac{s_2(x)}{s_1(x)}$.  Then \eqref{section} and
  condition~\ref{cs.1} together imply $a(x+y) = a(x)a(y)$, so
  $a\in G(\bFq)^*$.  The coboundary $\tau_0 : \Weil{\Fq} \to G(\bFq)^*$
  defined by $\tau_0(\Frob{\Fq}) \ceq \,^{\Frob{\Fq}} a \cdot a^{-1}$ satisfies $\tau_2 = \tau_0 \tau_1$,
  so $\tau_2$ and $\tau_1$ are cohomologous. Thus
  \begin{equation}\label{cohomologous}
  \begin{split}
    \obj \QC(G) &\to \Hh^1(\Weil{\Fq},G(\bFq)^*), \\
    \qcs{L} &\mapsto [\tau_\qcs{L}],
  \end{split}
  \end{equation}
  is independent of the choice of global sections made above.

  We now show that $[\tau_\qcs{L}]$ depends only on the isomorphism class of $\qcs{L}$.
  Let $\alpha : \qcs{L} \to \qcs{L'}$ be an isomorphism and
  let $\tau$ and $\tau'$ be any cocycles representing the classes in the
  image of $\qcs{L}$ and $\qcs{L'}$ under \eqref{cohomologous}; by construction,
  $\tau$ and $\tau'$ are defined by the choice of sections $s$ and $s'$.
  Now, $\varphi_x(s(x))$ is a
  non-zero scalar multiple of $s'(x)$; define $a: G(\bFq) \to \EEx$
  by $\alpha_x(s(x)) = a(x) s'(x)$.
  By the second part of condition~\ref{cs.3},
  $a(x+y) = a(x)a(y)$, so $a \in G(\bFq)^*$. By the first part of
  condition~\ref{cs.3}, $\tau'(\Frob{\Fq}) = \tau(\Frob{\Fq}) \cdot (\,^{\Frob{\Fq}^{-1}}a \cdot a^{-1})$,
  %\Clifton{Fix this.}
  so $\tau$ and $\tau'$ are cohomologous. This shows that
  \begin{equation}\label{pich1}
  \begin{split}
    \QCiso{G} &\to \Hh^1(\Weil{\Fq},G(\bFq)^*),\\
    [\qcs{L}] &\mapsto [\tau_\qcs{L}]
  \end{split}
  \end{equation}
  is well-defined.

@@

 To see that \eqref{pich1} is injective,
 suppose $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$.
 Let $s$ and $s'$ be sections appearing in the definition of $\tau_\qcs{L}$ and $\tau_\qcs{L'}$.
 Since $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$, 
 it follows that $\tau_\qcs{L}$ and $\tau_{\qcs{L'}}$ are cohomologous, 
 so there is some $a\in G(\Fq)^*$ such that 
 $\tau'(\Frob{\Fq}) = \tau(\Frob{\Fq}) \cdot (\,^{\Frob{\Fq}}a \cdot a^{-1})$. 
 Define $\alpha : \qcs{L}\to \qcs{L'}$ by $\alpha_x(s(x)) \ceq a(x) s'(x)$ and $\alpha^{-1} : \qcs{L}'\to \qcs{L}$
 by $\alpha^{-1}_x(s'(x)) \ceq a(x)^{-1} s(x)$, for $x\in G(\bFq)$.
 It is straightforward to check that $\alpha$ satisfies condition~\ref{cs.3}
 and that $\alpha^{-1}$ is its inverse. Thus $\qcs{L} \iso \qcs{L'}$ in $\QC(G)$ and \eqref{pich1} is injective.

 Finally, to show that \eqref{pich1} is surjective, pick $[\tau]\in \Hh^1(\Weil{\Fq}, G(\bFq)^*)$.
 For each $x\in G(\bFq)$, set $\gqcs{L}_x \ceq \EE$.
 Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by
 $\mu_{x,y}(1) \ceq 1\otimes1$, for $x,y\in G(\bFq)$; 
 clearly, $\mu$ satisfies condition~\ref{cs.1}.
 Define $\phi_x : \gqcs{L}_{\Frob{\Fq}(x)} \to \gqcs{L}_x$ by $\phi_x(1) \ceq \tau(\Frob{\Fq})(x)$,
 for each $x\in G(\bFq)$. 
 Then $\phi$ satisfies condition~\ref{cs.2} because $\tau \in Z^1(\Weil{\Fq},G(\bFq)^*)$.
 Setting $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$, we have that $[\tau]$ is the image of $\qcs{L}$
 under \eqref{pich1}, and thus \eqref{pich1} defines an isomorphism of groups.
\end{proof}

@@

Recall from Section~\ref{sec:finite} that 
a morphism $f : H\to G$ of group $\Fq$-schemes is a discrete isogeny 
if it is a finite, surjective \'etale morphism and 
the action of $\Gal(\bFq/\Fq)$ on the \'etale group scheme $\ker f$ is trivial.
